---
title: "Hypothesis Testing"
author: "Ahsan Ijaz"
date: "17/06/2016"
output:
  ioslides_presentation:
    keep_md: yes
    widescreen: yes
runtime: shiny
---

## Hypothesis Testing

- Null Hypothesis (Something that people expect)
- Alternate Hypothesis (An experiment conducted by you to challenge expectation)
- Hypothesis type (Z-test, t-test, ANOVA,...)
- Given values (Mean of population, variance of population,sample values, assumption)
- Critical Value or p-value (when to reject hypothesis)

## Normal Distribution

```{r, echo=FALSE}
library(ggplot2)
library(plyr)

mu01 = 500
sigma1 = 30
n1 = 30
xbar1 = 450
# dat1 <- data.frame(cond = factor(rep(c("Distribution of Panda Weight"), each=3000)), 
#                   rating = c(rnorm(3000,mean = mu01,sd = sigma1)))
# cdat1 <- ddply(dat1, "cond", summarise, rating.mean=mean(rating))

inputPanel(
  selectInput("sd_breaks", label = "Underlying Population:",
              choices = c(100, 200, 300, 400), selected = 200),
  
  sliderInput("bw_adjust", label = "Mean of alternate hypothesis",
              min = 0, max = 900, value = 500, step = 20),
  
    selectInput("crit_value", label = "P-value:",
              choices = c(0.005,0.05, 0.1, 0.2), selected =0.05)
)

renderPlot({
  set.seed(123)
  z_value = qnorm(p = (1-as.numeric(input$crit_value)),mean = mu01,sd = as.numeric(input$sd_breaks))
  z_value1 = qnorm(p = (1-as.numeric(input$crit_value)),mean = mu01,sd = as.numeric(input$sd_breaks),lower.tail = FALSE)
  dat1 <- data.frame(cond = factor(rep(c("Distribution"), each=3000)), 
                  rating = c(rnorm(3000,mean = mu01,sd = as.numeric(input$sd_breaks))))
cdat1 <- ddply(dat1, "cond", summarise, rating.mean=mean(rating))

ggplot(dat1, aes(x=rating, fill="blue")) + geom_density(alpha=.3) + geom_vline(data=cdat1, aes(xintercept=rating.mean,  colour=cond),linetype="dashed", size=1) + geom_vline(xintercept=input$bw_adjust) +
  geom_vline(xintercept=z_value,colour="red") +
  geom_vline(xintercept= z_value1,colour="red") + ggtitle(paste0("Area on left side is ",
                                                        pnorm(q= as.numeric(input$bw_adjust),mean = mu01,
                                                                               sd = as.numeric(input$sd_breaks))))

})

```


## Normal Distribution

```{r, echo=FALSE}

mu01 = 500
sigma1 = 30
n1 = 30
xbar1 = 450
# dat1 <- data.frame(cond = factor(rep(c("Distribution of Panda Weight"), each=3000)), 
#                   rating = c(rnorm(3000,mean = mu01,sd = sigma1)))
# cdat1 <- ddply(dat1, "cond", summarise, rating.mean=mean(rating))

inputPanel(
  selectInput("sd_breaks2", label = "Underlying Population:",
              choices = c(100, 200, 300, 400), selected = 200),
  
  sliderInput("bw_adjust2", label = "Mean of alternate hypothesis",
              min = 0, max = 900, value = 500, step = 20),
  
    selectInput("crit_value2", label = "P-value:",
              choices = c(0.005,0.05, 0.1, 0.2), selected =0.05)
)

renderPlot({
  set.seed(123)
  z_value = qnorm(p = (1-as.numeric(input$crit_value2)),mean = mu01,sd = as.numeric(input$sd_breaks2))
  z_value1 = qnorm(p = (1-as.numeric(input$crit_value2)),mean = mu01,sd = as.numeric(input$sd_breaks2),lower.tail = FALSE)
  dat2 <- data.frame(cond = factor(rep(c("Underlying Population","Sample Results"), each=3000)), 
                  rating = c(rnorm(3000,mean =as.numeric(input$bw_adjust2) ,sd=100),
                             rnorm(3000,mean = mu01,sd = as.numeric(input$sd_breaks2))))
cdat2 <- ddply(dat2, "cond", summarise, rating.mean=mean(rating))

ggplot(dat2, aes(x=rating, fill=cond)) + geom_density(alpha=.3) + geom_vline(data=cdat2, aes(xintercept=rating.mean,  colour=cond),linetype="dashed", size=1) + geom_vline(xintercept=input$bw_adjust2) +
  geom_vline(xintercept=z_value,colour="red") +
  geom_vline(xintercept= z_value1,colour="red") + ggtitle(paste0("Area on left side is ",
                                                        pnorm(q= as.numeric(input$bw_adjust2),mean = mu01,
                                                                               sd = as.numeric(input$sd_breaks2))))

})

```


## Lower tail test (known mean and variance)

Suppose the manufacturer claims that the mean lifetime of a battery is more than <b>500</b> hours. In a sample of <b>30</b> batteries, it was found that they only last <b>450</b> hours on average. Assume the population standard deviation is <b>12</b> hours. At <b>.05</b> significance level, can we reject the claim by the manufacturer? 

## Problem breakdown:

- Null hypothesis: mean is more than 500
- Alternate hypothesis: Mean is less than 500
- Hypothesis type: Z-test with lower tail
- Given values (mean and sd of population, mean of samples, number of samples)
- Critical value: 0.05

## Variables of Interest:

```{r}
mu02 = 500    
sigma2 = 12
n = 30
xbar = 450
z = (xbar-mu02)/(sigma2/sqrt(n))
z
```

## Visualization of problem:

```{r,echo = FALSE,fig.align='center'}

inputPanel(
  selectInput("sd_breaks1", label = "Population standard deviation:",
              choices = c(12, 50, 100, 200,1000), selected = 12),
  
  sliderInput("bw_adjust1", label = "Mean of alternate hypothesis",
              min = 350, max = 650, value = 450, step = 10)
)

renderPlot({
  set.seed(123)
#  z_value = qnorm(p = (1-as.numeric(input$crit_value)),mean = mu01,sd = as.numeric(input$sd_breaks))
#  z_value1 = qnorm(p = (1-as.numeric(input$crit_value)),mean = mu01,sd = as.numeric(input$sd_breaks),lower.tail = FALSE)
  dat3 <- data.frame(cond = factor(rep(c("Distribution"), each=3000)), 
                  rating = c(rnorm(3000,mean = mu02,sd = as.numeric(input$sd_breaks1))))
cdat3 <- ddply(dat3, "cond", summarise, rating.mean=mean(rating))
p_value = .05
z2 = (as.numeric(input$bw_adjust1-mu02))/(as.numeric(input$sd_breaks1)/sqrt(n))
z.alpha = qnorm(p_value,lower.tail=TRUE) ## 
if(z2 < z.alpha){
  print("Null hypothesis rejected, The mean time battery life is less than 500")
  varType = "Null hypothesis rejected"
}else{
  print("Cannot reject null hypothesis")
  varType = "Cannot reject null hypothesis"
}

ggplot(dat3, aes(x=rating, fill=cond)) + geom_density(alpha=.3) + geom_vline(data=cdat3, aes(xintercept=rating.mean,  colour=cond),linetype="dashed", size=1) + geom_vline(xintercept=input$bw_adjust1) + 
  ggtitle(varType)

})


# library(plyr)
# dat <- data.frame(cond = factor(rep(c("Distribution of batteries"), each=3000)), 
#                   rating = c(rnorm(3000,mean = mu0,sd = sigma)))
# cdat <- ddply(dat, "cond", summarise, rating.mean=mean(rating))
# ggplot(dat, aes(x=rating, fill=cond)) + geom_density(alpha=.3) + geom_vline(data=cdat, aes(xintercept=rating.mean,  colour=cond),
#                linetype="dashed", size=1) + geom_vline(xintercept=xbar) 
#+ geom_histogram(aes(y=..density..),binwidth=.5,
#                                                                                colour="black", fill="white")
```

## Test Hypothesis (lower tail):

```{r}

p_value = .05
z.alpha = qnorm(p_value,lower.tail=TRUE) ## 
if(z < z.alpha){
  print("Null hypothesis rejected, The mean time battery life is less than 500")
  varType = "Null hypothesis rejected"
}else{
  print("Cannot reject null hypothesis")
}
```

## Upper tail test (known mean and variance):

Suppose the food label on a cookie bag states that there is at most <b>2</b> grams of fat in a single cookie. In a sample of <b>35</b> cookies, it is found that the mean amount of saturated fat per cookie is <b>2.1</b> grams. The population standard deviation is <b>0.25</b> grams. At .05 significance level, can we reject the claim on food label?

## Problem breakdown:

- Null hypothesis: mean is less than 2
- Alternate hypothesis: Mean is greater than 2
- Hypothesis type: Z-test with upper tail
- Given values (mean and sd of population, mean of samples, number of samples)
- Critical value: 0.05

## Exploration of problem:

```{r}
mu0 = 2
sigma = 0.25
n = 35
xbar = 2.1
z = (xbar-mu0)/(sigma/sqrt(n))
z
```

## Visualization of problem:

```{r,echo = FALSE,fig.align='center'}
library(plyr)
dat <- data.frame(cond = factor(rep(c("Distribution of fat"), each=3000)), 
                  rating = c(rnorm(3000,mean = mu0,sd = sigma)))
cdat <- ddply(dat, "cond", summarise, rating.mean=mean(rating))
ggplot(dat, aes(x=rating, fill=cond)) + geom_density(alpha=.3) + geom_vline(data=cdat, aes(xintercept=rating.mean,  colour=cond),
               linetype="dashed", size=1) + geom_vline(xintercept=xbar)
#+ geom_histogram(aes(y=..density..),binwidth=.5,
#                                                                                colour="black", fill="white")
```

## Test hypothesis (upper tail): 

```{r}
p_value = .05
z.alpha = qnorm(p_value,lower.tail=FALSE)
if(z > z.alpha){
  print("Null hypothesis rejected, The food label is incorrect")
}else{
  print("The null hypothesis that food label is correct cannot be rejected.")
}
```

## Two tail test

Suppose the mean weight of Pandas last year was <b>120</b> kg. In a sample of <b>10</b> Pandas same time this year, the mean penguin weight is <b>114</b> kg. Assume the population standard deviation is <b>5.5</b> kg. At <b>.05</b> significance level, can we reject the null hypothesis that the mean penguin weight does not differ from last year?


## Problem breakdown:

- Null hypothesis: mean is equal to 120
- Alternate hypothesis: Mean is not equal to 120
- Hypothesis type: Z-test with both tails
- Given values (mean and sd of population, mean of samples, number of samples)
- Critical value: 0.05


## Exploration of problem:


```{r}
xbar =  114           # sample mean 
mu03 = 120            # hypothesized value 
sigma3 = 5.5            # population standard deviation 
n = 10                 # sample size 
z = (xbar-mu03)/(sigma3/sqrt(n))
z
```

## Visualization of problem:

```{r,echo = FALSE,fig.align='center'}
inputPanel(
  selectInput("sd_breaks3", label = "Population standard deviation:",
              choices = c(5.5, 10, 20, 50,1000), selected = 12),
  
  sliderInput("bw_adjust3", label = "Mean of alternate hypothesis",
              min = 50, max = 200, value = 114, step = 5)
)

renderPlot({
  set.seed(123)
#  z_value = qnorm(p = (1-as.numeric(input$crit_value)),mean = mu01,sd = as.numeric(input$sd_breaks))
#  z_value1 = qnorm(p = (1-as.numeric(input$crit_value)),mean = mu01,sd = as.numeric(input$sd_breaks),lower.tail = FALSE)
  dat4 <- data.frame(cond = factor(rep(c("Distribution"), each=3000)), 
                  rating = c(rnorm(3000,mean = mu03,sd = as.numeric(input$sd_breaks3))))
cdat4 <- ddply(dat4, "cond", summarise, rating.mean=mean(rating))
p_value = .05
z3 = (as.numeric(input$bw_adjust3-mu03))/(as.numeric(input$sd_breaks3)/sqrt(n))
z.alpha = qnorm(p_value,lower.tail=TRUE) ## 
if(z3 < z.alpha | z3 > -z.half.alpha){
  print("Null hypothesis rejected, The mean time battery life is less than 500")
  varType2 = "Null hypothesis rejected"
}else{
  print("Cannot reject null hypothesis")
  varType2 = "Cannot reject null hypothesis"
}

ggplot(dat4, aes(x=rating, fill=cond)) + geom_density(alpha=.3) + geom_vline(data=cdat4, aes(xintercept=rating.mean,  colour=cond),linetype="dashed", size=1) + geom_vline(xintercept=input$bw_adjust3) + 
  ggtitle(varType2)

})
# library(plyr)
# dat <- data.frame(cond = factor(rep(c("Distribution of Panda Weight"), each=3000)), 
#                   rating = c(rnorm(3000,mean = mu0,sd = sigma)))
# cdat <- ddply(dat, "cond", summarise, rating.mean=mean(rating))
# ggplot(dat, aes(x=rating, fill=cond)) + geom_density(alpha=.3) + geom_vline(data=cdat, aes(xintercept=rating.mean,  colour=cond),
#                linetype="dashed", size=1) + geom_vline(xintercept=xbar)
#+ geom_histogram(aes(y=..density..),binwidth=.5,
#                                                                                colour="black", fill="white")
```

## Test hypothesis (two tail):

```{r}
alpha = .05
z.half.alpha = qnorm(alpha/2)
if(z < z.half.alpha | z > -z.half.alpha){
  print("Null hypothesis rejected, The weight of Pandas haven't changed")
}else{
  print("Cannot reject the null hypothesis.")
}
```

# T-distribution:

## When do we need it

- So far we knew the underlying standard deviation of the population. 
- In case we don't, we need to somehow estimate the standard deviation by looking at our data
- More samples we have of the data, more confident we are about the normality of distribution 

## T-distribution visualization

```{r, echo=FALSE}
library(ggplot2)
library(plyr)


inputPanel(
  
  sliderInput("dof", label = "Degree of freedom",
              min = 5, max = 100, value = 10, step = 5)
)

renderPlot({
  set.seed(123)
  dat3 <- data.frame(cond = factor(rep(c("Normal Distribution","T-distribution"), each=10000)), 
                  rating = c(rnorm(10000),rt(n =10000,df = as.numeric(input$dof))))

ggplot(dat3, aes(x=rating,colour = cond)) + geom_density(alpha=.3) 
})

```

## One sample T-test:

An outbreak of Dengue was attributed to water collected in drums at a
certain factory. Scientists measured the level of Dengue in 9 randomly sampled
water drums. The levels were:
0.593 0.142 0.329 0.691 0.231 0.793 0.519 0.392 0.418
Is there evidence that the mean level of virus in the drums is greater than 0.3?

## Problem breakdown:

- Null hypothesis: mean is equal to 0.3
- Alternate hypothesis: Mean is greater than 0.3
- Hypothesis type: t-test one sample
- Given values (values of samples collected)
- Critical value: 0.05

## Hypothesis Test (one sample)

```{r}
x = c(0.593, 0.142, 0.329, 0.691, 0.231, 0.793, 0.519, 0.392, 0.418)
t.test(x, alternative="greater", mu=0.3)
```

## Hypothesis Test (two samples)

6 subjects were given a drug (treatment group) and an additional 6 subjects a
placebo (control group). Their reaction time to a stimulus was measured (in ms). We
want to perform a two-sample t-test for comparing the means of the treatment and
control groups.

## Two sample T-test:
Let muMed be the mean of the population taking medicine and muPlacebo the mean of the untreated
population. Here the hypothesis of interest can be expressed as:
- H0 => muMed - MuPlacebo = 0
- Ha => muMed - MuPlacebo > 0
- Hypothesis type: t-test two samples
- Given values (values of samples collected)
- Critical value: 0.05

```{r}
Control = c(91, 87, 99, 77, 88, 91)
Treat = c(101, 110, 103, 93, 99, 104)
t.test(Control,Treat,alternative="greater", var.equal=TRUE)
```

## Hypothesis test (paired)
A study was performed to test whether cars get better mileage on premium gas than
on regular gas. Each of 10 cars was first filled with either regular or premium gas,
and the mileage for that tank was recorded. The mileage was
recorded again for the same cars using the other kind of gasoline. We need to use a paired ttest
to determine whether cars get significantly better mileage with premium gas.


## Two sample T-test:

- H0 => Premium is similar to regular gas for mileage
- Ha => Premium > Mileage
- Hypothesis type: t-test paired
- Given values (values of samples collected)
- Critical value: 0.05


## Paired T-tests:

```{r}
reg = c(16, 20, 21, 22, 23, 22, 27, 25, 27, 28)
prem = c(19, 22, 24, 24, 25, 25, 26, 26, 28, 32)
t.test(prem,reg,alternative="greater", paired=TRUE)
```


